\_sh v3.0  701  IrulaPhonemicTextU

\id 1.LitWorkshop

\ref LitWkshp.1 (1-4 Angel)
\tx ni           sʌrɨ tindelɐ?
\mb ninu         sʌrɨ *tind -ɐlle
\ge you          rice   ***   -instrumental
\ps pro          n      ***   -case
\mbtam நினு ***    ***   -***

\ft Have you eaten (rice)?


\ref LitWkshp.2
\tx nɐmu         coːɾɾu          t̪ind̪emu
\mb nɐːmu       coːɾɾu          t̪in        -t̪t̪    -eːmu
\ge we           food               eat          -past      -1pl
\ps pro          n                  v            -asp       -s.agr
\mbtam நாம் சோர்ரு தின் -த்த -எமு

\ft We ate food.


\ref LitWkshp.3
\tx niʋiɾi coːɾɾu          t̪ind̪iɾijɐ?
\mb niʋiɾi coːɾɾu          t̪in        -t̪t̪    -iɾi  -ɐ
\ge 2pl      food               eat          -past      -2pl   -ques
\ps pro      n                  v            -asp       -s.agr -ques
\mbtam ***      சோர்ரு தின் -த்த -***   -***

\ft Have you eaten?


\ref LitWkshp.4
\tx nɐmɨmu                  sʌrɨ tɪndemu.
\mb nɐmu         - *ɨ -mu   sʌrɨ tɪnd -emu
\ge we           - *** -and  rice   eat   -1pl.past
\ps pro          - *** -conj n      v     -s.agr
\mbtam நாம் - *** -***  ***    ***   -***

\ft We also have eaten (rice).

\id 2.RoomSentences

\ref JPDN60.01
\tx ceʋuɾu           ʋeɭɭeːnd̪u            iɾukud̪u.
\mb ceʋuɾu           ʋeɭɭeː         -t̪u   iɾu      -kk   -ud̪u
\ge wall               be.white           -ptc    being     -pres -inanim
\ps n                  v                  -v.suff v         -asp  -s.agr.png
\mbtam செவுரு வெள்ளே -து இரு -***  -து

\ft The wall is white.


\ref JPDN60.02
\tx ceʋuɾud̪i               pɐɖɐː   ʋɐɾen̪d̪uɾukkɐːɾu
\mb ceʋuɾu           -t̪i   pɐɖɐː   *ʋɐɾen̪d̪uɾukkɐːɾu
\ge wall               -in     picture   ***
\ps n                  -pp     n         ***
\mbtam செவுரு -தி படா ***

\ft A drawing is on the wall.


\ref JPDN60.03
\tx ɐgge            pɐɭɭineː             kuːʈʈugɐ.
\mb ɐgge            pɐɭɭi         -neː   kuːʈʈu          -kk   -ɐɭ
\ge Mother          floor           -ACC    sweep              -pres -3sf
\ps n               n               -case   v                  -asp  -s.agr
\mbtam அக்கெ பள்ளி -னே கூட்டி -***  -அள

\ft Mother is sweeping the floor.


\ref JPDN60.04
\tx pɐɭɭike                       pɐːjine             ʋiɾɯt̪u        kukki
\mb pɐɭɭi         - ke            pɐːji       -ne     ʋiɾɯ      -t̪u kukki
\ge floor           - to.allative   mat          -ACC    put          -GEN  sit
\ps n               - case          n            -case   v            -case v
\mbtam பள்ளி - -க்கே பாயி -னெ விரு -***  குக்

\ft Put the mat on the floor and sit.


\ref JPDN60.05
\tx ɐgge            kɐd̪ɐːʋune        t̪orɐ
\mb ɐgge            kɐd̪ɐːʋu  -ne     t̪orɐ
\ge Mother          door         -ACC    open
\ps n               n            -case   v
\mbtam அக்கெ கதவு -னெ தொற

\ft Mother, open the door.


\ref JPDN61.06
\tx ɐmmi            dʒɐnɐːlune            t̪oɖeʈʈu             ʋe
\mb ɐmmi            dʒɐnɐːlu      -ne     t̪oɖeʈʈu             *ʋe
\ge sister          window          -ACC    wipe.clean               ***
\ps n               n               -case   v                        ***
\mbtam அம்மி ஜனாலு -னெ தொடெத்து ***

\ft Sister, clean the window.


\ref JPDN61.07
\tx kɐʈʈɐku                     meːle       kukkɐ                  ʋɐɳɖɐː
\mb kɐʈʈɐ         -ku           meːle       kukki        -ɐ        ʋɐːɳɖɐː
\ge table           -DAT          top          sit          -3sf.past dont.want
\ps n               -case         pos          v            -s.agr    v
\mbtam கட்டா -க்கு மேலெ குக் -***      வண்டா

\ft Don't sit on the table.


\ref JPDN61.08
\tx ceɾɾune          t̪ukki                    konɖu               ʋɐː
\mb ceːɾu      -neː tuːkku            -t̪u   koɭɭu      -t̪u   ʋɐː
\ge chair        -ACC  take.up            -ptc    hold         -ptc    come
\ps n            -case v                  -v.suff v            -v.suff v
\mbtam சேறு -ஏ  தூக்கு -து கொள் -து வா

\ft Bring the chair.


\ref JPDN61.09
\tx mɐd̪ilukku                   meːle       eːɾɐ            ʋɐɳɖɐː
\mb mɐd̪ilu        -kku          meːle       eːɾu -kkɐ       ʋɐːnɖɐː
\ge shelf           -from.DAT     top          climb  -inf       don't.want
\ps n               -case         pos          v      -v.suff    v
\mbtam மதிலு -க்கு மேலெ ஏற -க்க வாண்டா

\ft Don't climb over the shelf.


\ref JPDN61.10
\tx ceʋuɾukke                     ɐːɳine          ɐɖikkɐ           ʋɐɳɖɐː
\mb ceʋuɾu           -keː        ɐːɳi    -ne     ɐɖi      -kkɐ    ʋɐɳɖɐː
\ge wall               -to.allative nail      -ACC    apply     -inf    don't
\ps n                  -case        n         -case   v         -v.suff v
\mbtam செவுரு -கே      ஆணி -னெ ஆடி -க் வண்டா

\ft Don't nail on the wall.


\ref JPDN61.11
\tx t̪oɾɐːppukɐːjeʈut̪t̪u kuːɾeːne          puːʈʈu
\mb t̪oɾɐːppukɐːje -t̪u     kuːɾeː    -ne     puːʈʈu
\ge lock               -GEN      house        -ACC    lock
\ps n                  -case     n            -case   v
\mbtam ***                -***      கூரே -னெ பூட்டு

\ft Lock the house with the lock.


\ref JPDN62.12
\tx ellɐːɾu ceːnd̪u          poːʈʈɐːʋu           eɖukkilɐː
\mb ellɐːɾu ceːn̪d̪u        poːʈʈɐːʋu           eɖu      -kk   -ilɐː
\ge everyone  together           photo                    take.up   -pres -potential
\ps n         adv                n                        v         -asp  -v.suff
\mbtam ***       சேந்து போட்டாவு எடு -***  -இலா

\ft All of us should take a photo together.


\ref JPDN62.13
\tx kuːɾeːke               poːji       puʈʈu            t̪iŋgilɐː                     ʋɐːji
\mb kuːɾeː    -keː        poː   - *ji puʈʈu            t̪in        -kkilɐː            ʋɐn +   ji
\ge house        -to.allative go     - *** food               eat          -hortative          come *** ***
\ps n            -case        v      - *** n                  v            -mod                v    *** ***
\mbtam கூரே -கே      போ - *** புட்டு தின் -க்கிலா ***  *** ***

\ft Let us go to the house and eat, come.


\ref JPDN62.14
\tx mɐːrukke                  cut̪t̪ɐmɐː         kuːɾeːne          kuːʈʈu
\mb mɐːru       -kke          cut̪t̪ɐmɐː         kuːɾeː    -ne     kuːʈʈu
\ge broom        -INSTRUM      clean                 house        -ACC    sweep
\ps n            -case         adj                   n            -case   v
\mbtam மாறு -க்கெ சுத்தமா கூரே -னெ கூட்டு

\ft Sweep the house clean with a broom.

\id 3.ForestSentences

\ref JPDN66.01
\tx mɐɾɐːt̪t̪e       ʋoʈʈɐː         ʋɐːɳɖɐː
\mb mɐɾɐː   - *t̪t̪e ʋoʈʈɐː         ʋɐɳɖɐː
\ge tree      - ***      cut                don't
\ps n         - ***      v                  v
\mbtam மரா - ***      வொட்டா வண்டா

\ft Don't cut the tree.


\ref JPDN66.02
\tx bod̪eːkkulu                    mɐcɐːlu        iɾukkud̪u
\mb bod̪eː     -kkulu              mɐcɐːlu        iɾu      -kud̪u
\ge bush         -in                 rabbit          being     -pres.3si
\ps n            -pp                 n               v         -s.agr
\mbtam பொதே -க்குலு மசாலு இரு -து

\ft Rabbit is in the bush.


\ref JPDN66.03
\tx keɾeːd̪i          ɐːɳeː  t̪ɐɳɳi       kuɖikkud̪u
\mb keɾeː      -d̪i   ɐːɳeː  t̪ɐɳɳi       kuɖi        -kk   -ud̪u
\ge river        -LOC.in elephant  water           drink        -pres -inanim
\ps n            -case   n         n               v            -asp  -s.agr
\mbtam கெரே -தி ஆணே தண்ணி குடி -***  -து

\ft Elephant is drinking water in the river.


\ref JPDN66.04
\tx kujine                     mɐccɐ        t̪oːɳɖugæ.
\mb kuji               -neː   mɐccɐ        t̪oːɳɖu        -kk   -ɐn
\ge hole               -ACC    cousin       dig                -pres -3sm
\ps n                  -case   n            v                  -asp  -s.agr
\mbtam குய்யி -னே மச்ச தோண்டு -***  -அன்

\ft Cousin is digging the pit.


\ref JPDN66.05
\tx pɐːreːd̪i          t̪unigɐ             kɐːgud̪u
\mb pɐːreː      -t̪i   t̪uni       -gɐ     kɐː   -kud̪u
\ge rock          -in     cloth        -plural dry    -pres.3si
\ps n             -pp     n            -n.suff v      -s.agr
\mbtam பாறே: -தி துனி -க    கா -குது

\ft Clothes are dried on the rock.


\ref JPDN67.06
\tx kɐllune               t̪ukki                    jeːɾɐːkke           ʋe
\mb kɐllu           -neː tuːkku            -t̪u   jeːɾɐː -kke         ʋe
\ge stone           -ACC  take.up            -ptc    house     -on.allative keep
\ps n               -case v                  -v.suff pos       -case        v
\mbtam கல்லு -ஏ  தூக்கு -து ***       -கெ      ***

\ft Take the stone and keep it inside.


\ref JPDN67.07
\tx t̪ɐɖineː       eɖut̪t̪u       nɐːjine             muɖukku
\mb t̪ɐɖi   -neː   eɖu      -t̪u   nɐːji       -neː   muɖukku
\ge stick     -ACC    take      -ptc    dog          -ACC    force
\ps n         -case   v         -v.suff n            -case   v
\mbtam தடி -னே எடு -து நாயி -னே முடுக்கு

\ft Take the stick and beat the dog.


\ref JPDN67.08
\tx puʋʋu            ceɖid̪i            ɐɭɐgɐː     iɾukkud̪u
\mb puːʋu            ceɖi        -d̪i   ɐɭɐgɐː     iɾu      -kk   -ud̪u
\ge flower             plant        -LOC.in beautiful    be        -pres -inanim
\ps n                  n            -case   adj          v         -asp  -s.agr
\mbtam புவ்வு செடி -தி அழகா இரு -***  -து

\ft Flower in the plant is beautiful.


\ref JPDN67.09
\tx jeleneː             puː   t̪ingud̪u
\mb jele         -ne     puː   t̪in        -kud̪u
\ge leaf         -ACC    worm   eat          -pres.3si
\ps n            -case   n      v            -s.agr
\mbtam யெலெ -னெ பூ தின் -குது

\ft Worm is eating the leaves.


\ref JPDN67.10
\tx ʋeːɾoːɖe        ceɖɨne             puɖungi                 ʋɐː
\mb ʋeːɾ   -oːɖe    ceɖɨ       -ne     puɖungi                 ʋɐː
\ge roots     -with      plant        -ACC    pluck                    come
\ps n         -case      n            -case   v                        v
\mbtam வேர -ஒடெ செடு -னெ புடுங்கி ***

\ft Pluck the plant with roots.


\ref JPDN68.11
\tx koɖi         bod̪eːd̪i         pɐɖɐːnd̪uɾukkud̪u
\mb koɖi         bod̪eː     -t̪i   pɐɖɐː -t̪u   -iɾu      -kk   -ud̪u
\ge creeper plant bush         -in     lay     -ptc    -perfect   -pres -inanim
\ps n             n            -pp     v       -v.suff -v.aux     -asp  -s.agr
\mbtam கொடி  பொதே -தி பட  -து -இரு -***  -து

\ft Creeper plant is on the bush.


\ref JPDN68.12
\tx me     ʋɐnd̪ɐːkki       kuːgɐː     moɭeːkkud̪u
\mb me     ʋɐn - ɐ-    *ːkki kuːgɐː     moɭeː      -kud̪u
\ge rain   come - that- ***    mushroom     grow         -pres.3si
\ps n      v    - det-  ***    n            v            -s.agr
\mbtam மெ ***  - அ-  ***    கூகா மொழே -குது

\ft Mushroom grows when it rains.

\id 4.BodyPartsSentences

\ref JPDN78.01
\tx mukkud̪i                  mɐccɐ        iɾukkud̪u.
\mb mukku              -t̪i   mɐccɐ        iɾu      -kku          -d̪u
\ge nose               -in     mole         being     -inf          -pres.3si
\ps n                  -pp     n            v         -v.suff       -s.agr
\mbtam முக்கு -தி மச்ச இரு -க்கு -து

\ft There is a mole on the nose.


\ref JPDN78.02
\tx net̪t̪ine                ceʋurukkeː                 iɖut̪t̪ut̪t̪eː.
\mb net̪t̪i          -neː   ceʋuɾu           -kkeː    iɖut̪t̪u -t̪t̪    -eː
\ge forehead           -ACC    wall               -on.ALLAT hit         -past      -1sg
\ps n                  -case   n                  -case     v           -asp       -s.agr
\mbtam நெத்தி -னே செவுரு -கே   இடு   -த்த -ஏ

\ft I hit the forehead on the wall.


\ref JPDN78.03
\tx meguru             kɐreːnd̪u             iɾukkud̪u.
\mb meguru             kɐreː    -n̪d̪u      iɾu      -kku          -d̪u
\ge hair               black     -adj          being     -inf          -pres.3si
\ps n                  adj       -adj          v         -v.suff       -s.agr
\mbtam மெகுறு கரே -ந்து இரு -க்கு -து

\ft The hair is black.


\ref JPDN78.04
\tx kɐːd̪u     bellɐd̪ɐː                iɾukkud̪u.
\mb kɐːd̪u     bellɐ              -d̪ɐː iɾu      -kku          -d̪u
\ge ear          big                -emph   being     -inf          -pres.3si
\ps n            adj                -emph   v         -v.suff       -s.agr
\mbtam காது பெல்லா -தா இரு -க்கு -து

\ft The ear is big.
Note: could use this as examples for [a:]



\ref JPDN78.05
\tx mɐt̪t̪ud̪i           iɾɐt̪t̪ɐ         ʋɐɾukkud̪u.
\mb mɐt̪t̪u       -d̪i   iː  - ɾɐt̪t̪ɐ   ʋɐː  -kk   -ud̪u
\ge mouth           -LOC.on this - blood        come   -pres -inanim
\ps n               -case   det  - n            v      -asp  -s.agr
\mbtam மத்து -தி இ  - றத்த வா -***  -து

\ft It is bleeding on the mouth.


\ref JPDN78.06
\tx pɐːd̪d̪ɐt̪t̪i      ʋeɖippu                ʋɐnd̪d̪ud̪u.
\mb pɐːd̪d̪ɐ     -t̪i   ʋeɖippu                ʋɐn -kku          -d̪u
\ge foot            -in     crack                    come -inf          -pres.3si
\ps n               -pp     n                        v    -v.suff       -s.agr
\mbtam பாத்த -தி வெடிப்பு ***  -க்கு -து

\ft There is a crack on the foot.


\ref JPDN79.07
\tx ʋɐɾɐlu     nɨʈʈɐmɐː          iɾukkud̪u.
\mb ʋɐɾɐlu     nɨʈʈɐmɐː          iɾu      -kku          -d̪u
\ge finger       long                  being     -inf          -pres.3si
\ps n            adj                   v         -v.suff       -s.agr
\mbtam வரலு நீட்டமா இரு -க்கு -து

\ft The finger is long.


\ref JPDN79.08
\tx kʌji        murund̪ud̪u.
\mb kʌji        muru         -n̪d̪u      -t̪u
\ge hand         broken       -adj          -ptc
\ps n            v            -adj          -v.suff
\mbtam கெயி முறு -ந்து -து

\ft The hand is broken.


\ref JPDN79.09
\tx keɳɖe-kɐːlu                  oɾu     ʋɐnijɐː iɾukkud̪u.
\mb keɳɖe-kɐːlu                  oɾu     ʋɐnijɐː iɾu      -kk   -ud̪u
\ge leg.calf                        a        pain      be        -pres -inanim
\ps n                               det      n         v         -asp  -s.agr
\mbtam கெண்டை காலு ஒரு***       இரு -***  -து

\ft The leg calf hurts.


\ref JPDN79.10
\tx t̪oɖed̪i          ennukku                    kɐːjɐː     ɐːjit̪t̪ud̪u.
\mb t̪oɖe      -t̪i   enn          -ku           kɐːjɐː     ɐːi      -t̪t̪    -ud̪u
\ge thigh        -in     1s           -DAT          wound        happen    -past      -inanim
\ps n            -pp     pro          -case         n            v         -asp       -s.agr.png
\mbtam தொடெ -தி என்ன -க்கு காயா ஆய் -த்த -து

\ft There is a wound in the thigh.


\ref JPDN79.11
\tx nɐgɐ   ennukku                    dʒɐctijɐː              ʋɐɭɐɾugud̪u.
\mb nɐgɐ   enn          -ku           dʒɐctijɐː              ʋɐɭɐ       -kk   -ud̪u
\ge n      1s           -DAT          more                     grow         -pres -inanim
\ps nail   pro          -case         adv                      v            -asp  -s.agr.png
\mbtam நக என்ன -க்கு ஜாஸ்டியா வளரு -***  -து

\ft Nail is growing more for me.

\id 5.ElicitngGrammar

\ref JGDN15.10
\tx ɐd̪u     emut̪u              kuːɾeː,   nit̪t̪ɐ            ɐllɐ.
\mb ɐd̪u     nɐːmu       -ud̪u  kuːɾeː    ninu         -ud̪u  ɐllɐ
\ge that      we           -GEN    house        you          -GEN    not
\ps dem       pro          -case   n            pro          -case   neg
\mbtam அது நாம் -து கூரே நினு -து அல்ல

\ft That's our house, not yours.

\ref JGDN16.09
\tx ellɐ         ponɖiɾigɐ                iɾukɐɾu.
\mb ellɐ         ponɖu             -gɐɭ   iɾu      -kk   -ɐɾu
\ge all          lady               -plural be        -pres -3pl
\ps quant        n                  -n.suff v         -asp  -s.agr
\mbtam எல்ல பெண்டு -க    இரு -***  -அரு

\ft There are all ladies.


\ref JGDN28.19
\tx keɖekke        ponɐɭɐ?
\mb keɖe -kkeː    poː   -t̪t̪    -ɐɭ    -ɐ
\ge get   -to.ALLAT go     -past      -3sf    -ques
\ps v     -case     v      -asp       -s.agr  -ques
\mbtam ***   -கே   போ -த்த -அள -***

\ft Did she go to the river?


\ref JPDN52.6
\tx ellɐ         ʋelene              muɖit̪ud̪u.
\mb ellɐ         ʋeːle      -neː   muɖi        -t̪t̪    -ud̪u
\ge all          work         -ACC    finish       -past      -inanim
\ps quant        n            -case   v            -asp       -s.agr
\mbtam எல்ல வேலெ -னே முடி -த்த -து

\ft All the work is finished.
